A Tale of Two Transects

### 2019-06-05 writing/group_presentation.Rnw

<<setup, include=FALSE>>=
knitr::opts_chunk$set(echo = FALSE, device = "tikz")
library(drake)
library(magrittr)
library(tidyverse)
library(ggplot2)
library(scales)
library(ggtree)
library(castor)
theme_set(theme_bw(base_size = 24) + theme(legend.direction = "vertical"))
loadd(bc_corr, bc_corr2, unif_corr, unif_corr2, qstats_join, datasets, longtree,
      taxon_ITS2_unite)
reads <- qstats_join %>%
  mutate_at("Region", factor,
            levels = c("ITS2", "ITS1", "short", "ITS", "LSU", "long")) %>%
  mutate_at("Dataset", factor,
            levels = c("short-ion", "short-pacbio", "long-pacbio")) %>%
  group_by(Dataset, Plate, Region)
reads_subset <- reads %>%
  do(if (nrow(.) > 10000) sample_n(., 10000) else .)
reverselog_trans <- function(base = exp(1)) {
    trans <- function(x) -log(x, base)
    inv <- function(x) base^(-x)
    trans_new(paste0("reverselog-", format(base)), trans, inv, 
              log_breaks(base = base), 
              domain = c(1e-100, Inf))
}
@

@@

\begin{frame}{Tree}
\framesubtitle{Kingdoms}
<<kingtree, fig.width = 12, message=FALSE>>=
seqhash <- function(seq, len = 8) UseMethod("seqhash")
seqhash.character <- function(seq, len = 8) {
  h <- purrr::map_chr(seq, digest::digest)
  stringr::str_sub(h, end = len)
}
seqhash.XStringSet <- function(seq, len = 8) {
  seqhash.character(as.character(seq))
}
treetax <- taxon_ITS2_unite %>%
  mutate_at("seq", seqhash) %>%
  select(-Taxonomy, -nreads) 
outgroups <- treetax %>% filter(!is.na(Kingdom), Kingdom %in% c("Alveolata", "Chromista", "Plantae"), seq %in% longtree$tip.label)
p <- ggtree(tr = longtree)
longtree <- reroot(longtree, MRCA(p, outgroups$seq))

>>>>

<<setup, include=FALSE>>=
knitr::opts_chunk$set(echo = FALSE, device = "tikz")
library(drake)
library(magrittr)
library(tidyverse)
library(ggplot2)
library(scales)
library(ggtree)
library(castor)
theme_set(theme_bw(base_size = 24) + theme(legend.direction = "vertical"))
loadd(bc_corr, bc_corr2, unif_corr, unif_corr2, qstats_join, datasets, longtree,
      taxon_LSUcons_rdp)
reads <- qstats_join %>%
  mutate_at("Region", factor,
            levels = c("ITS2", "ITS1", "short", "ITS", "LSU", "long")) %>%
  mutate_at("Dataset", factor,
            levels = c("short-ion", "short-pacbio", "long-pacbio")) %>%
  group_by(Dataset, Plate, Region)
reads_subset <- reads %>%
  do(if (nrow(.) > 10000) sample_n(., 10000) else .)
reverselog_trans <- function(base = exp(1)) {
    trans <- function(x) -log(x, base)
    inv <- function(x) base^(-x)
    trans_new(paste0("reverselog-", format(base)), trans, inv, 
              log_breaks(base = base), 
              domain = c(1e-100, Inf))
}
@

@@

\begin{frame}{Tree}
\framesubtitle{Kingdoms}
<<kingtree, fig.width = 12, message=FALSE>>=
seqhash <- function(seq, len = 8) UseMethod("seqhash")
seqhash.character <- function(seq, len = 8) {
  h <- purrr::map_chr(seq, digest::digest)
  stringr::str_sub(h, end = len)
}
seqhash.XStringSet <- function(seq, len = 8) {
  seqhash.character(as.character(seq))
}
treetax <- taxon_LSUcons_rdp %>%
  mutate_at("seq", seqhash) %>%
  select(-Taxonomy, -nreads) 
outgroups <- treetax %>% filter(!is.na(Kingdom), Kingdom %in% c("Alveolata", "Chromista", "Plantae"), seq %in% longtree$tip.label)
p <- ggtree(tr = longtree)
longtree <- reroot(longtree, MRCA(p, outgroups$seq))

### 2019-06-05 writing/transect_paper.Rmd

>>>>

## Sampling

  - Ouémé Forest reserve
  - Two locations in _Isoberlinia doka_ forest, approx. 30km apart
  - transect: 25m, sampled at interval of 1m
  - Sampled in May 2015 and June 2016
  - remove coarse debris (typically very little) and take ~5cm×5cm×5cm sample
  - homogenize by hand in ziplock bag
  - take $\approx$ 50mg, lysis in field with Terralyzer
  - remainder of sample dried at 40°C with electric dryer within 24-48 hours

## Soil Chemistry

  - Need to get methods from Nourou

## DNA extraction, PCR, and sequencing

  - lysis in field Zymo Research Xpedition kit
  - short amplicon : gITS7 -- ITS4 (multiplexing index on gITS7)
  - long amplicon : ITS1 -- LR5 (multiplexing index on both ends)
  - short amplicon : same library sequenced with PacBio RS II and Ion Proton S5
  - long amplicon : PacBio RS II

## Bioinformatics

  - Two workflows for short reads:
    - Mohammad's workflow in Pipecraft
    - Brendan's workflow using Snakemake/Drake

  - Mohammad's pipeline:

  - Brendan's pipeline:
      - Ion Torrent
          - used demultiplexing from SciLife (standard IonXpress)
          - primers removed with `cutadapt` [@martin2011]
      - PacBio
          - CCS basecalls re-done with `ccs` version ___ [@pacificbiosciences2019]
          - demultiplexed and primers removed with `cutadapt` [@martin2011]
      - ITS1, ITS2, ITS, LSU extracted using ITSx [@Nilsson2010; @bengtsson-palme2013]
      - extracted regions independently quality filtered by max expected error using `dada2::filterAndTrim()`
      - dereplicated using `dada2::derepFastq()`
      - denoised with `dada2::dada()`
      - de novo chimera removal using `dada2::removeBimeraDenovo()`
      - taxonomy assigned using IDTAXA [@murali2018]
          - UNITE version 8 [@nilsson2019a]
              - extracted ITS1 and ITS2 using ITSx [@Nilsson2010; @bengtsson-palme2013]
              - trained three classifiers: full ITS, ITS1, ITS2.
          - RDP fungal LSU release 11.5 [@cole2014]
              - chose only sequences containing conserved primer sites for ITS4 (3' side of H11; CAUAUnAnUnAGnSSAGGA) and LR5 (on H37; CGAAnUUUCnCUCAGGAUAGCnG).
              - ambiguities based on [@cannone2002; @tedersoo2015]
              - Trim all bases after LR5 because they are not present in our amplicons.
          - SILVA LSU release 132 [@quast2013; @glockner2017; @yilmaz2014]
              - filter out sequences annotates as prokaryotes, then treat as RDP.
          - all databases
              - select medioid sequence for each annotated species, based on minimum 5-mer distance to other sequences
              - for sequences not annotated as species (e.g. "unidentified Basidiomycota"), randomly choose up to 20 per taxon
              - Use up to 5 iterations of IDTAXA training, removing all "problem" sequences on each iteration, unless last member of annotated taxon.
              - "underannotated" sequences (e.g., "unidentified Basidiomycota" confidently placed as Agaricomycetes) not protected from removal if last sequence annotated as such.
      - ecological guild assigned using FUNGuild [@nguyen2016funguild]
      - consensus long read and LSU region constructed for each ITS2 ASV:
          - align reads using `DECIPHER` with RNA secondary structure calculation [@wright2015;@wright2016]
          - remove outliers (i.e. chimeras) with `odseq` [@jehl2015]
          - remove gap columns (>50%)
          - take consensus (>50%) at non-gaps
          - remove sequences with more than 3 ambiguous bases (bad consensus)
      - initial alignment of long amplicons with DECIPHER [@wright2015;@wright2016]
      - alignment/tree for long amplicons built iteratively with PASTA[@mirarab2014; @mirarab2014a]. Fastree tree, mafft align, OPAL merge, RaxML final tree

## Statistics
